---
title: "Week 4 Assignment"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE, warning = TRUE)

library(ottr)
library(testthat)
```

---

## Overview

---

This assignment covers the content in week 4 lectures and tutorials.

#### Mark breakdown
- Question 1 (9 marks)
- Question 2 (7 marks)
- Question 3 (4 marks)

---

## Background

---

BioID is a method for determining which proteins are in close proximity to one another within the cell. In this assignment we will work with data from the Human Cell Map: (ref)

To ensure that the main data set is entered correctly, a matrix with the SAINT scores for each protein pair is provided. A SAINT score is measurement of the confidence that 2 proteins interact with one another. In the matrix below, a score of 1 indicates that 2 proteins definitely interact, and 0 indicates that there is no evidence of interaction. Along the diagonal (proteins vs. themselves) all values are 1.

```{r}
PPI_scores <- matrix(c(1,0,0,1,0,0,1,0,0,0,0,0,0,1,0,1,0,0,1,1,0,0,0.98,0,0,0,0,0,0,0,1,0,0,0,0,1,0,0,0,0,1,0.78,1,0,0,1,0,0,1,0,0,0,0,0,0,1,0,1,0,0,1,1,0,0,1,0.99,0,0.3,0,0,0,1,0,0,1,1,0,0,1,1,0.95,1,0,0,1,0,0,1,0,0,1,0,0,0,0,0,0,1,0,0,1,0,0,0,0,1,0,0,0,0,0,0,0,0.98,0,0,1,1,0,0,1,0,0,1,0,0,0,0,0,0,0.99,1,0,0,0,1,1,0,0,0,0,0,0,0,0,0.95,0,0,0,1,1,0,0,0,0,0,0,0,0.3,1,0,0,1,0,0,1,0,0,0,0,1,0,0,0,0,0,0,0,0,0,1,0.5,1,0,0.78,1,0,0,1,0,0,0,0,0,0.5,1), nrow=14)
proteins <- c("AARS2","CCDC14","COIL","CS","PCM1","PCNT","PDHA1","PRPF8","SASS6","SEPT10","SEPT9","STIL","TERF2IP","TRAP1")
rownames(PPI_scores) <- proteins
colnames(PPI_scores) <- proteins
PPI_scores
```

Looking at the above matrix you can see that AARS1 interacts with CS, and that AARS1 does not interact with COIL.

---

### Question 1 

Each of these proteins is localized to a specific cellular compartment. The table below shows the cellular compartment and subcompartment for 12 of the 14 proteins.

| Protein | Compartment  | Subcompartment |
| ------- | ------------ | -------------- |
| AARS2   | mitochondria | matrix         |
| CCDC14  | centrosome   | -              |
| COIL    | nucleus      | nuclear body   |
| CS      | mitochondria | matrix         |
| PCM1    | centrosome   | -              |
| PCNT    | centrosome   | -              |
| PRPF8   | nucleus      | nuclear body   |
| SEPT10  | cytoskeleton | actin          |
| SEPT9   | cytoskeleton | actin          |
| STIL    | centrosome   | -              |
| TERF2IP | nucleus      | nuclear body   |
| TRAP1   | mitochondria | matrix         |

#### a. 
(3 marks)
Create a character matrix called "cellular_localization" with 12 rows and 2 columns. Row names should be the protein names, and the two columns should be named "compartment" and "subcompartment". Ensure you copy the names of the proteins and column names exactly. You can create the matrix with any method.
```{r Assignment.W4.1.a}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W4.1.a.R")
```
**IGNORE the above block: test results will not appear in assignments** 

#### b.
(2 marks) 
Two proteins were not included in the above matrix as they have an unknown cellular localization. Add two new rows to the matrix with the information below.

| Protein | Compartment  | Subcompartment |
| ------- | ------------ | -------------- |
| SASS6   | ?            | ?              |
| PDHA1   | ?            | ?              |
```{r Assignment.W4.1.b}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W4.1.b.R")
```
**IGNORE the above block: test results will not appear in assignments** 

#### c.
(1 mark) 
Use **named** indexing to output the subcompartment for the protein PRPF8. Assign the value to a variable named "PRPF8_subcompartment".
```{r Assignment.W4.1.c}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W4.1.c.R")
```
**IGNORE the above block: test results will not appear in assignments** 

#### d.
(1 mark)
Use **positional** indexing to output the main compartment for all proteins. Assign the output vector to a variable named "all_main_compartments".
```{r Assignment.W4.1.d}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W4.1.d.R")
```
**IGNORE the above block: test results will not appear in assignments** 

#### e.
(1 mark)
Use logical indexing to output all of the rows (proteins) in `cellular_localization` with a main compartment equal to "mitochondria". Assign the output matrix to a variable named "mitochondrial_proteins".
```{r Assignment.W4.1.e}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W4.1.e.R")
```
**IGNORE the above block: test results will not appear in assignments** 

#### f.
(1 mark)
Use logical indexing to output all of the rows (proteins) in `cellular_localization` with a main compartment equal to "centrosome". Assign the output matrix to a variable named "centrosome_proteins".
```{r Assignment.W4.1.f}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W4.1.f.R")
```
**IGNORE the above block: test results will not appear in assignments** 

---

### Question 2

#### a.
(3 marks)
Create a list called "PPI_data" with the following elements in the order they are listed. Choose the appropriate data type and structure for each. The final two are the matrices already created in the assignment. Do not include names for any of the list elements.

- "HEK293"
- "Go, et al. A proximity-dependent biotinylation map of a human cell. Nature (2021)"
- 14
- `cellular_localization`
- `PPI_scores`
```{r Assignment.W4.2.a}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W4.2.a.R")
```
**IGNORE the above block: test results will not appear in assignments** 

#### b.
(1 mark)
Use **positional** indexing to output the value of the second element in `PPI_data`. Assign the output value to a variable named "PPI_data_element_2".
```{r Assignment.W4.2.b}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W4.2.b.R")
```
**IGNORE the above block: test results will not appear in assignments** 

#### c.
(1 mark)
Name the list elements in `PPI_data`. Use the names in the order specified below. 

- "cell_line"
- "reference"
- "num_proteins"
- "cellular_localization"
- "PPI_scores"
```{r Assignment.W4.2.c}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W4.2.c.R")
```
**IGNORE the above block: test results will not appear in assignments** 

#### d.
(1 mark)
Use **named** indexing to output the full cellular localization information for the protein AARS2 (both the main compartment and subcompartment) from `PPI_data`. Assign the output vector to a variable named "AARS2_localization".
```{r Assignment.W4.2.d}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W4.2.d.R")
```
**IGNORE the above block: test results will not appear in assignments** 

#### e.
(1 mark)
Add a new element called "proteins" to the `PPI_data` list. It should be a character vector with the names of all 14 proteins in alphabetical order. (Hint: you should check the row names or the column names of `PPI_scores`.)
```{r Assignment.W4.2.e}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W4.2.e.R")
```
**IGNORE the above block: test results will not appear in assignments** 

---

### Question 3

The cellular localization for two of the proteins (SASS6 and PDHA1) were not provided. In this question we will determine which compartments each protein likely belongs to.

#### a.
(1 mark)
Load the pheatmap package.
```{r Assignment.W4.3.a}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W4.3.a.R")
```
**IGNORE the above block: test results will not appear in assignments** 

#### b.
(1 mark)
Create a heat map of the list element `PPI_scores`. Assign the call to the `pheatmap()` function to a variable named "PPI_heatmap". It will still create a plot.
```{r}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W4.3.b.R")
```
**IGNORE the above block: test results will not appear in assignments** 

#### c.
(1 mark)
Locate PDHA1 on the heat map and identify the proteins with which it definitely interacts (score of 1). Look back at the output from question 1.e. and 1.f.
Which main compartment does the protein PDHA1 likely localize to? 
Assign the name of one of the 5 main compartments to a variable named "PDHA1_main_compartment".
```{r Assignment.W4.3.c}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W4.3.c.R")
```
**IGNORE the above block: test results will not appear in assignments** 

#### d.
(1 mark)
Locate SASS6 on the heat map and identify the proteins with which it definitely interacts (score of 1). Look back at the output from question 1.e. and 1.f. 
Which main compartment does the protein SASS6 likely localize to? 
Assign the name of one of the 5 main compartments to a variable named "SASS6_main_compartment".
```{r Assignment.W4.3.d}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W4.3.d.R")
```
**IGNORE the above block: test results will not appear in assignments** 

---