---
title: "4.1 Introduction to Matrices - Tutorial"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE, warning = TRUE)

library(ottr)
library(testthat)
```
**Run the above block to load required packages**

---

At the end of this tutorial you should be able to:

* Create matrices using the matrix function
* Add rows and column names to matrices
* Index matrices with positional indexing
* Index matrices with named indexing
* Add new rows and columns to matrices
* Alter row and column names

---

## 4.1.1 Matrix Creation

---

Create a matrix containing the numbers 1 to 15 in 3 rows, such that the first row contains the numbers: 1, 4, 7, 10, and 13. Assign it to a variable named "matrix_1".

```{r}

```

```{r}
. = ottr::check("tests/Tutorial.W4.1.1.a.R")
```
**Check your answer by running the above block** 

Create a matrix containing the numbers 1 to 15 in 3 rows, such that the rightmost number in the first row is 5. Assign it to a variable named "matrix_2".

```{r}

```

```{r}
. = ottr::check("tests/Tutorial.W4.1.1.b.R")
```
**Check your answer by running the above block** 

---

For the rest of this tutorial we will focus on data for the frequency of single nucleotide polymorphisms (SNPs) in different populations. SNP frequencies from the 1000 Genomes Project are sourced from dbSNP (https://www.ncbi.nlm.nih.gov/snp/).

We will start by looking at a few randomly chosen SNPs across 4 populations:
- African (AFR)
- East Asian (EA)
- European (EUR)
- South Asian (SA)

SNPs have ids in the form "rsXXXX" where the Xs are an ID number. Here we will look at the SNPs rs671, rs4680, and rs6265.

These are the frequencies of the *alternative allele* in each of the four populations:

|     | rs671 | rs4680 | rs6265 |
| --- | ----- | ------ | ------ |
| AFR | 0.00  | 0.28   | 0.01   |
| EA  | 0.17  | 0.28   | 0.49   |
| EUR | 0.00  | 0.50   | 0.20   |
| SA  | 0.00  | 0.44   | 0.20   |

The following vector contains the values in the table above, going from left to right along each row in the table:

```{r}
SNP_frequencies <- c(0.00, 0.28, 0.01, 0.17, 0.28, 0.49, 0.00, 0.50, 0.20, 0.00, 0.44, 0.20)
```

Use the `matrix()` function to create a matrix that matches the table above using the `SNP_frequencies` vector. Assign the matrix to a variable named "SNPs". Verify that the order of the numbers in your matrix matches the table!

```{r}

```

```{r}
. = ottr::check("tests/Tutorial.W4.1.1.c.R")
```
**Check your answer by running the above block** 

Add row names to `SNPs`.

```{r}

```

```{r}
. = ottr::check("tests/Tutorial.W4.1.1.d.R")
```
**Check your answer by running the above block** 

Add column names to `SNPs`.

```{r}

```

```{r}
. = ottr::check("tests/Tutorial.W4.1.1.e.R")
```
**Check your answer by running the above block** 

---

## 4.1.2 Basic Matrix Indexing

---

Using positional indexing, assign the value in the 3rd row, 2nd column in `SNPs` to a variable named "SNP_3_2".

```{r}

```

```{r}
. = ottr::check("tests/Tutorial.W4.1.2.a.R")
```
**Check your answer by running the above block** 

Using positional indexing, assign the 2nd row in `SNPs` to a variable named "SNPs_row_2".

```{r}

```

```{r}
. = ottr::check("tests/Tutorial.W4.1.2.b.R")
```
**Check your answer by running the above block** 

Using named indexing, assign the alternate allele frequencies for rs671 and rs6265 in the EA and EUR populations to a variable named "SNPs_subset".

```{r}

```

```{r}
. = ottr::check("tests/Tutorial.W4.1.2.c.R")
```
**Check your answer by running the above block** 

Using individual matrix element indices, index:
- the value in the 2nd row, 1st column
- the value in the 3rd row, 2nd column
- the value in the 1st row, 3rd column

Assign the output vector to a variable named "SNP_indexed_elements".

```{r}

```

```{r}
. = ottr::check("tests/Tutorial.W4.1.2.d.R")
```
**Check your answer by running the above block** 

---

## 4.1.3 Matrix Modification

---

Add a new column of alternative allele frequencies for the SNP rs7412 to the `SNPs` matrix:

|     | rs7412 |
| --- | ------ |
| AFR | 0.10   |
| EA  | 0.10   |
| EUR | 0.06   |
| SA  | 0.04   |

```{r}

```

```{r}
. = ottr::check("tests/Tutorial.W4.1.3.a.R")
```
**Check your answer by running the above block** 

Add a new row for the population: American (AMR)

|     | rs671 | rs4680 | rs6265 | rs7412 |
| --- | ----- | ------ | ------ | ------ |
| AMR | 0.00  | 0.32   | 0.15   | 0.05   |

```{r}

```

```{r}
. = ottr::check("tests/Tutorial.W4.1.3.b.R")
```
**Check your answer by running the above block** 

Update the row names so that all the names are three letters long by changing "EA" to "EAS" and "SA" to "SAS".

```{r}

```

```{r}
. = ottr::check("tests/Tutorial.W4.1.3.c.R")
```
**Check your answer by running the above block** 

---