---
title: "4.3 Lists"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE, warning = TRUE)
```

---

## 4.3.1 List Creation & Indexing

---

### Lists

Vectors and matrices only allow one data type to be stored in them. However, often we have many different types of data to store. 

For example, if we were saving data about a particular patient, we may want to save all of the following pieces of information:

Patient ID: PID0012839   
Age: 57    
Sex: F    
Insured: TRUE
Appointments: 

|          | 21/10/2018 | 07/11/2019 | 27/03/2020 | 02/08/2020 |
| -------- | ---------- | ---------- | ---------- | ---------- |
| Check-up | TRUE       | TRUE       | FALSE      | FALSE      |

Blood Tests:

|            | HGB  | RBC | WBC | BG |
| ---------- | ---- | --- | --- | -- |
| 21/10/2018 | 12.3 | 5.2 | 6.4 | 76 |
| 07/11/2019 | 13.4 | 5.3 | 6.8 | 82 |
| 02/08/2020 | 13.5 | 5.2 | 6.6 | 79 |

*Lists* are an R data structure that solves this problem. A list is a collection of elements that can be different data types and/or data structures.

Lists are 1-dimensional like vectors, however each element in a list does not have to be a single value like a vector. 

List elements can be:
- single values (vectors of length 1)
- vectors
- matrices
- lists 

Let's create a list called "pt_info" to hold the following patient information:

Patient ID: PID0012839   
Age: 57    
Sex: F    
Insured: TRUE

```{r}
pt_info <- list("PID0012839", 57, "F", TRUE)
pt_info
```

Looking at the output we can see that there are four elements, and each element is preceded by a number X in double square brackets: `[[X]]`

---

### List Indexing

Lists are indexed using double square brackets:
    
    list_name[[index]]
    
For example, if we wanted to retrieve the patient age that is at index 2, we would write:

```{r}
pt_info[[2]]
```

You can see that the output looks like a regular vector of length 1. If we only used single brackets it would look different:

```{r}
pt_info[2]
```

This time, we have retrieved the data at the second index, but it has been returned as a list of length 1. 

It's important to be mindful when indexing lists as this can lead to issues. For example, if we want to add 1 to the patient's age, this will work:

```{r}
pt_info[[2]] + 1
```

But this will not:

```{r}
pt_info[2] + 1
```

---

### List Element Names

Lists can also have names, which can be viewed, added, or updated with the `names()` function.

Let's add names to the elements in `pt_info`:

```{r}
names(pt_info) <- c("pt_ID", "age", "sex", "insured")
pt_info
```

Names can also be used as indices for lists and must be put into double square brackets to access the element itself stored at that position:

```{r}
pt_info[["insured"]]
```

Again, indexing with single square brackets will return a list of length 1:

```{r}
pt_info["insured"]
```

Lists have a second option for named indexing. If you look at the output of `pt_info["insured"]` above, you will see that each element, once named, is preceded by a dollar sign and the element name, for example `$insured`. Individual elements of a list can be accessed with the following syntax:

    list_name$element_name
    
For example:

```{r}
pt_info$insured
```

```{r}
pt_info$pt_ID
```

This type of indexing will only work if:
- the elements have names
- the names do NOT have spaces

---

## 4.3.2 Working with Lists

---

### Complex Lists

Earlier this week we learned the `matrix()` function but ignored one of the arguments: `dimnames`. The `dimnames` argument takes a list of length 2 with row and column names for the matrix you are creating.

Let's try using this argument while creating a matrix to store the patient blood test data: 

|            | HGB  | RBC | WBC | BG |
| ---------- | ---- | --- | --- | -- |
| 21/10/2018 | 12.3 | 5.2 | 6.4 | 76 |
| 07/11/2019 | 13.4 | 5.3 | 6.8 | 82 |
| 02/08/2020 | 13.5 | 5.2 | 6.6 | 79 |

First let's create a list of row and column names:

```{r}
dates <- c("d.21.10.2018", "d.07.11.2019", "d.02.08.2020")
measurements <- c("HGB", "RBC", "WBC", "BG")
blood_dim_names <- list(dates, measurements)
blood_dim_names
```

The variable `blood_dim_names` has a length of 2, where the first element is a vector of row names and the second element is a vector of column names. This list can now be used in the creation of a matrix named "blood" to hold the blood measurement data:

```{r}
blood <- matrix(c(12.3, 13.4, 13.5, 5.2, 5.3, 5.2, 6.4, 6.8, 6.6, 76, 82, 79), 
                nrow = 3, 
                dimnames = blood_dim_names)
blood
```

Let's also create a vector to hold information about whether each appointment the patient had was a check-up:

|          | 21/10/2018 | 07/11/2019 | 27/03/2020 | 02/08/2020 |
| -------- | ---------- | ---------- | ---------- | ---------- |
| Check-up | TRUE       | TRUE       | FALSE      | FALSE      |

```{r}
appt_dates <- c("d.21.10.2018", 
                "d.07.11.2019", 
                "d.27.03.2020", 
                "d.02.08.2020")
appt_is_checkup <- c(TRUE, TRUE, FALSE, 
                     FALSE)
names(appt_is_checkup) <- appt_dates
appt_is_checkup
```

Now we can create another list to hold all of the patient information. Note that names can be assigned to list elements during creation, as we do here:

```{r}
patient <- list("patient_info" = pt_info,  
                "appointments" = appt_dates,                     
                "checkup" = appt_is_checkup, 
                "blood_tests" = blood)
patient
```

The list `patient` we created contains: 
- a list (with names)
- a character vector
- a logical vector (with names)
- a numeric matrix (with names)

To access elements within the list `patient`, often chained indexing is required. For example, to access the sex of the patient we need to first access the list element `patient_info` and then access the element `sex` within that list. The indices can be chained together. First, we index `patient` with the `patient_info` index, then we index `patient_info` with the `sex` index:

```{r}
patient$patient_info$sex
```

Similarly, if we want to access a column in the element `blood_tests`, we index `patient` with the `blood_tests` index, and then we can index the matrix.

For example, we can retrieve the column "HGB" from the `blood_tests` element:

```{r}
patient$blood_tests[, "HGB"]
```

If we wanted to retrieve the second appointment data from the `appointments` element, we can access the vector and then the second vector element:

```{r}
patient$appointments[2]
```

---

### Editing Lists

At two of the patient's appointments, they received a referral to another doctor.

|	         | Doctor   | Specialty  | Reason             |
| ---------- | -------- | ---------- | ------------------ |
| 07/11/2019 | Jay, J.  | ENT        | Tinnitus           |
| 02/08/2020 | Chen, D. | Cardiology | Heart palpitations |

Let's create a matrix to hold this information:

```{r}
referral_names <- list(c("d.27.03.2020", "d.02.08.2020"),
                       c("Doctor", "Specialty", "Reason"))
referral_matrix <- matrix(c("Jay, J.", "Chen, D.", "ENT", "Cardiology", 
                            "Tinnitus", "Heart palpitations"), 
                          nrow = 2, dimnames = referral_names)
referral_matrix
```

We can add this new matrix to the `patient` list by assigning it to a named list element that does not yet exist. There are two options for this, as there are two ways to access named list elements:

        list_name$new_name <- new_element
        list_name[["new_name"]] <- new_element

```{r}
patient$referrals <- referral_matrix
patient$referrals
```

Let's create another list for all of the patient's doctors. The patient's GP is H. Davies, and the first referral the patient had was to J. Jay:

```{r}
doctors <- list("Davies, H.", "Jay, J.")
doctors
```

After the patient's most recent referral, they have a new doctor: D. Chen. To add the new doctor to the doctor list (which has elements with no names), we need to assign the character value to a new positional index. 

Looking at the output of `doctors` above you can see that there are currently two elements, and the next element will have a positional index of 3. The best way to consistently choose the correct positional index for adding a new element is to take the current length of the list and add one:

        list_name[[length(list_name) + 1]] <- new_element
        
```{r}
doctors[[length(doctors) + 1]] <- "Chen, D."
doctors
```

If we had assigned "Chen, D." to `doctors[[1]]` or `doctors[[2]]`, we would've overwrite the existing elements. If we had assigned "Chen, D." to `doctors[[4]]`, it would work, but it would have also added a new empty element to `doctors[[3]]`.

To change an individual element, as mentioned, it can be overwritten by assigning a new value to the list at the index of interest:

        list_name[[existing_index]] <- new_element
        
For example, if Dr. H. Davies was actually Dr. H. Davis, we can fix it by assigning "Davis, H." to `doctors[[1]]`:

```{r}
doctors[[1]] <- "Davis, H."
doctors
```

---

## 4.3.3 Using Lists in Bioinformatic Analysis

---

### The strsplit() Function

Some functions return lists. One example is the `strsplit()` function, which splits a character value into separate substrings (components of the original character value). How the character value is split is determined by the *delimiter*, which is a character(s) that denotes separate "chunks" of data. The result of splitting a character value on a delimiter is a vector containing shorter "chunks" of the original value, where each chunk (also known as a *substring*) is a separate element in the vector, and the delimiter is no longer present. The `strsplit()` function returns the vector as part of a list.

Usage: `strsplit(x, split)`
Arguments: `x` - character value or vector of character values
           `split` - character value to define delimiter at which to split
Return value: list of vectors, one vector per character value in `x`

For example, if we want to split the character value "Davis, H." at the comma, the comma would be our delimiter:

```{r}
strsplit("Davis, H.", split = ",")
```

It returns a list of length 1, which contains a vector of length 2. Splitting the original character value at the comma resulted in two "chunks": one character value with all of the letters before the comma, and one with all the letters after. 

Another useful function when working with lists is `unlist()`. The `unlist()` function turns a list into a vector, but can only be used if all of list elements are of the same type, such as those in the doctors list (character type):

```{r}
doctors_vector <- unlist(doctors)
doctors_vector
```

Now we can split all three character values in `doctors_vector` at once with `strsplit()`, splitting at the comma:

```{r}
strsplit(doctors_vector, split = ",")
```

Now it returns a list of length 3, with a character vector at each element. Note that the same delimiter applies to all elements of the input vector.

---

### Calculating GC Content

The `strsplit()` function can be very helpful when working with DNA, RNA, or amino acid sequences. If the `split` argument is set to an empty character value "", it will split the character values at each letter:

```{r}
DNAsequences <- c("ACTGTCGTTAGCAGT", 
                  "TATGCCGTATCGTTAG")
split_seqs <- strsplit(DNAsequences, 
                       split = "")
split_seqs
```

Each DNA sequence is now a vector of bases in the `split_seqs` list.

Now we can perform analyses like calculating the GC content of each sequence. First we can check which bases are "G" or "C" in the first sequence:

```{r}
split_seqs[[1]] == "G" | split_seqs[[1]] == "C"
```

Another helpful function is the `which()` function. The `which()` function takes a logical vector and returns the positional indices of the elements which are `TRUE`. It answers the question "which indices are `TRUE`?":

```{r}
which(split_seqs[[1]] == "G" |  split_seqs[[1]] == "C")
```

We can then determine the length of the vector returned by `which()`:

```{r}
numGC <- length(which(split_seqs[[1]] == "G" | split_seqs[[1]] == "C"))
numGC
```

Now we can determine the total number of bases in the first sequence and calculate the proportion of the sequence that is G or C:

```{r}
numBases <- length(split_seqs[[1]])
numGC/numBases
```

We can nest all of this together and directly calculate the GC content in one line of code. Let's use the second sequence as an example:

```{r}
length(which(split_seqs[[2]] == "G" | split_seqs[[2]] == "C")) / length(split_seqs[[2]])
```

---